---
title: "Illustrative Examples"
author: "Andrew Fullerton"
date: "2024-11-07"
output: github_document
---

```{r}
library(sqldf)
```

```{r}
mortality_ill <- read.csv("data/m3_mortality.csv")
```

# SQL: Creating and accessing dataframes in R

```{r}
str(mortality_ill)
```

### Run this code. What happened?

```{r warning=TRUE}
sqldf(
  "CREATE TABLE mortality_ill_small AS
   SELECT uid,
          death_date,
          sex
   FROM mortality_ill"
)
```

### Now run this code. What happened?

```{r}
mortality_ill_small <- sqldf(
  "SELECT uid,
          death_date,
          sex
   FROM mortality_ill"
)

str(mortality_ill_small) # You should see that your R object was created
```

This is because SQL is not integrated into R via `sqldf` the same way that `PROC SQL` integrates SQL with SAS. We cannot write data within SQL queries passed into `sqldf`. 

# SQL: Accessing native R functions in `sqldf`

### Run this code. What happened?

```{r error=TRUE}
mortality_ill_small_v2 <- sqldf(
  "SELECT uid,
          as.Date(death_date, format = '%m/%d/%Y')
   FROM mortality_ill"
)
```

### Now run this code. What happened?

```{r}
mortality_ill_small_v2 <- sqldf(
  "SELECT uid,
          death_date
   FROM mortality_ill"
)

mortality_ill_small_v2$death_date <- as.Date(mortality_ill_small_v2$death_date,
                                             format = '%Y-%m-%d')

str(mortality_ill_small_v2) # Let's see the new dataset we made
```

Our SQL query is not integrated into the R language the same way that `PROC SQL` integrates SQL queries into SAS. We cannot (easily) access or use R functions in SQL queries passed to `sqldf`.

We can only use standard SQL functions in `sqldf`.

# GIS: Reading shapefiles into our environment

Try to find a function that will read this shapefile into R. Don't load any packages! Hint: most base R read functions begin with `read`.

```{r}
filepath <- "shapefile/lha.shp" # Try to read this filepath in the space below

# ...
# ...
# ...
```

Any luck?

# GIS: Try to find base R functions for spatial data

Now ask your favourite generative AI tool or search engine. What'd you find? Note that ChatGPT may try to (incorrectly) tell you that R has a package included by default called 'maps' that works with spatial data.


